---
title: "Create your own Website with R Markdown"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Learn how to create a website with R Markdown.
- Create a personal or project website.
- Chapter 1 and Section 3.1 of [blogdown: Creating Websites with R Markdown](https://bookdown.org/yihui/blogdown/)

# Motivation

- **You should have a personal website**
    - Makes you stand out compared to a LinkedIn page that only boring HR folks
      seem to care about.
    - You can post links to class projects and research projects so future
      employers can tell that you're legit.
    - My website is a little outdated (I made it back in grad school), but
      it has still helped me during my career: <https://dcgerard.github.io/>
  
- You can create a project website:
    - This facilitates open-source collaboration between researchers.
    - E.g. I used such a website for notekeeping and collaborating with
      my PI during my postdoc: <http://dcgerard.github.io/succotash_sims/analysis/index.html>

- Other uses:
    - Advertise specific software/papers <https://dcgerard.github.io/updog/>.
    - Create a blog (say, on data science projects) <https://dcgerard.github.io/research/2018/03/16/updog.html>.
    
# Installation

- We'll use the blogdown R package to build a website, so make sure you have
  it installed:
  
    ```{r, eval = FALSE}
    install.packages("blogdown")
    ```

- [Hugo](https://gohugo.io/) is a software that builds a website for you. 
  Blogdown uses Hugo to make clean-looking websites. You can install it via:
    ```{r, eval = FALSE}
    blogdown::install_hugo()
    ```

# Getting started

1. Set your working directory to the location on your local machine
   where you want to save your website. Do this via 
   
    > Session > Set Working Directory > Choose Directory...

2. Run the following in your R consul:

    ```{r, eval = FALSE}
    blogdown::new_site()
    ```

- This will populate a bunch of folders containing the contents of your website.

- The `public/` folder contains the contents of your website. You will be able to upload everything in the `public/` folder to a host to publish your website online.

- The other folders contain files used to make the files

- You can view your website via 

    > Addins > Serve Site
    
    A window with a preview of the website will pop up.
    
# Editing your website

## New Theme

## config.toml

## content/

- This folder is where you place R Markdown files. These R Markdown files will be converted to HTML files.

# Publishing your website
