---
title: "Simple Features and Graphing Maps"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

# Learning Objectives

- Understand the concept and utility of spatial data visualization in R.
- Get acquainted with the basic functions and data types in the `{sf}` package.
- Create simple maps using the `{sf}` package.
- Simple Features
  - <https://r-spatial.github.io/sf/articles/>
- Map Plotting
  - <https://ggplot2.tidyverse.org/reference/ggsf.html>

# Motivation

- For some reason, people are really impressed with data visualizations that include maps (e.g. of D.C., the U.S., the Earth, etc).

- Let's go over how to build simple map visualizations in R.

- For this lecture, we will try to visualize the 2020 presidential election elections from (<https://github.com/TheUpshot/presidential-precinct-map-2020>).

# Introduction to Simple Features in R

-   The `sf` package, implements the "simple features" standard, making it easier to manipulate geographical data in R.

    ```{r}
    library(sf)
    ```

