---
title: "Reactivity and the server() Function"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Learn the basics reactivity and the the server() function in Shiny Apps.
- Chapter 4 of [Mastering Shiny](https://mastering-shiny.org/).
    - Most content discretely stolen from Hadley's book.
- [Shiny Cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/shiny.pdf).
- Optional Resources
  - [Shiny Tutorial](https://stat545.com/shiny00_index.html).

# The Server Function

## Render Fuctions

- You've already seen render functions. They are of the form:
    ```{r, eval = FALSE}
    render*({
      ## Code goes here.
    })
    ```
    
- The curly braces allow you to write multiple lines of code and submit that
  as the argument.
    
- Always save output of render function as an element of the `output` list.

- Any `input` elements inside the `render*()` function will trigger
  the render code to be re-evaluated when those `input` elements change.
