---
title: "Function Creation"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Learning Objectives

- Creating your own functions.
- Chapter 2 of [HOPR](https://rstudio-education.github.io/hopr/)
- Chapter 26 of [RDS](https://r4ds.hadley.nz/)

## Functions

- All functions are of the form

    ``` r
    name <- function(arg1, arg2 = default1, arg3 = default2) {
      ## Code using arg1, arg2, arg3, to create result
      return(result)
    }
    ```

- You can choose any `name` you want, but they should be informative.
- You choose the names of the arguments `arg1`, `arg2`, etc...
    - These are the inputs the user will use.
- Arguments can have defaults by setting `arg1 = default1`, where `default1` is whatever the default value of arg1 is. In the above example, `arg1` has no default but `arg2` and `arg3` have defaults.
- Your code creates some output which I call `result` above.
- You put the output in a `return()` call at the end of the function.

- Steps to creating a function:

    1. figure out the logic in a simple case
    2. name it something meaningful - usually a verb
    3. list the inputs inside function(x,y,z)
    4. place code for function in a {} block 
    5. test your function with some different inputs
    6. add error-checking of inputs
    
- Coding standards

    1. Save as text file
    2. Indent code
    3. Limit width of code (80 columns?)
    4. Limit the length of individual functions
    5. Frequent comments
    
    ```{r}
    add_two <- function(a, b) {
      return(a + b)
    }
    add_two(2, 4)
    ```
    

- Example from our book follows.
    ```{r bookexample_rescale}
    df <- tibble::tibble(
      a = rnorm(10),
      b = rnorm(10),
      c = rnorm(10),
      d = rnorm(10)
    )
    
    df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
      (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
    df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
      (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
    df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
      (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
    df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
      (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
    ```

- How many inputs does each line have?
    ```{r redo1}
    x <- df$a
    (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
    # get rid of duplication
    rng <- range(x, na.rm = TRUE)
    (x - rng[1]) / (rng[2] - rng[1])
    # make it into a function and test it
    rescale01 <- function(x) {
      rng <- range(x, na.rm = TRUE)
      (x - rng[1]) / (rng[2] - rng[1])
    }
    rescale01(c(0, 5, 10))
    rescale01(c(-10, 0, 10))
    rescale01(c(1, 2, 3, NA, 5))
    df$a <- rescale01(df$a)
    df$b <- rescale01(df$b)
    df$c <- rescale01(df$c)
    df$d <- rescale01(df$d)
    ```

- Now, if we have a change in requirements, we only have to change it in one place.  For instance, perhaps we want to handle columns that have Inf as one of the values.

    ```{r rescale_with_Inf}
    x <- c(1:10, Inf)
    rescale01(x)
    rescale01 <- function(x) {
      rng <- range(x, na.rm = TRUE, finite = TRUE)
      (x - rng[1]) / (rng[2] - rng[1])
    }
    rescale01(x)
    ```

- Do's and do not's of function naming:

    * pick either snake_case or camelCase but don't use both
    * meaningful names (preferably verbs)
    * for a family of functions, start with the same word
    * try not to overwrite common functions or variables
    * use lots of comments in your code, particularly to explain the "why"
      of your code or to break up your code into sections using something 
      like `# load data --------------------`
      
- **Exercise**: Write a function that calculates the $z$-scores of a numeric vector. The $z$-score takes each value, subtracts the mean, then divides the standard deviation. It is the measure of how many standard deviations above (or below) the mean a value is.

- **Exercise**: Write a function that takes a numeric vector as input and replaces all instances of `-9` with `NA`.

- **Exercise**: Write a function that takes a numeric vector and returns the coefficient of variation (the mean divided by the standard deviation).

- **Exercise**: Write a function that takes as input a vector and returns the number of missing values.

- **Exercise** (from RDS): Given a vector of birth dates, write a function to compute the age in years.

    ```{r, echo = FALSE, eval = FALSE}
    age <- function(x) {
      (x %--% lubridate::today()) / years(1)
    }
    ```

- **Exercise**: Re-write the the function `range()`. Use functions: `min()`, `max()`

    ```{r, echo = FALSE, eval = FALSE}
    # Returns the minimum and maximum elements of a numeric vector.
    # 
    # x: A numeric vector
    #
    # returns: A vector length 2. The first element is the minimum of x and the
    #          second element is the maximum of x.
    range2 <- function(x) {
      return(c(min(x), max(x)))
    }
    ```


- **Exercise**: Write `both_na()`, a function that takes two vectors of the 
  same length and returns the number of positions that have an `NA` in both 
  vectors. Useful functions: `is.na()`, `sum()`, 
  logical operators.
  
    ```{r, eval = FALSE, echo = FALSE}
    # Returns number of positions that have an NA in both positions of two vectors.
    #
    # x: A vector the same length as y
    # y: A vector the same length as x
    #
    # returns: The number of positions that have an NA in both x and y
    both_na <- function(x, y) {
      stopifnot(length(x) == length(y))
      return(sum(is.na(x) & is.na(y)))
    }
    both_na(c(1, NA, 2), c(NA, NA, 2))
    both_na(c(NA, NA, 2), c(NA, NA, 2))
    both_na(c(NA, NA, NA), c(NA, NA, 2))
    ```

- **Exercise**: Read the source code for each of the following three functions, 
  describe what they do, and then brainstorm better names.

    ```{r, eval = FALSE}
    f1 <- function(string, prefix) {
      substr(string, 1, nchar(prefix)) == prefix
    }
    
    f2 <- function(x) {
      if (length(x) <= 1) return(NULL)
      x[-length(x)]
    }
    
    f3 <- function(x, y) {
      rep(y, length.out = length(x))
    }
    ```
    
    ```{r, eval = FALSE, echo = FALSE}
    # Checks if string contains a given prefix
    #
    # string: A string.
    # prefix: Another string.
    #
    # returns: TRUE if the first characters in string match prefix and FALSE
    #          otherwise.
    matches_prefix <- function(string, prefix) {
      substr(string, 1, nchar(prefix)) == prefix
    }
    
    # Removes the least element of a vector
    # 
    # x: A vector
    #
    # returns: A vector. Equal to x but with the last element removed.
    pop <- function(x) {
      if (length(x) <= 1) return(NULL)
      x[-length(x)]
    }
    
    # cycles through the elements of y until it has the same length as x
    #
    # y: A vector to be repeated
    # x: A vector whose length is to be matched.
    #
    # returns: A vector with the elements of y repeated until it matches the 
    #          length of x.
    stupid_rep <- function(x, y) {
      rep(y, length.out = length(x))
    }
    ```
