---
title: "Function Creation"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

## Functions

- When to write a function and why to write a function

- Steps to creating a function:

    1. figure out the logic in a simple case
    2. name it something meaningful - usually a verb
    3. list the inputs inside function(x,y,z)
    4. place code for function in a {} block 
    5. test your function with some different inputs
    6. add error-checking of inputs
    
- Coding standards

    1. Indent code
    2. Limit width of code (80 columns?)
    3. Limit the length of individual functions
    4. Frequent comments
    
    ```{r}
    add_two <- function(a, b) {
      return(a + b)
    }
    add_two(2, 4)
    ```



