---
title: "Style Guide"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
    toc_float: false
    highlight: pygments
urlcolor: "blue"
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, 
                      fig.width  = 4)
knitr::opts_knit$set(global.par = TRUE)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
par(mar = c(3, 3, 2, 1),
mgp = c(1.8, 0.4, 0), 
las = 1,
tcl = -.25)
```

# Learning Objectives

- Coding Style
- [tidyverse style guide](https://style.tidyverse.org/)
- [Google style guide](https://google.github.io/styleguide/Rguide.html)
- [Bioconductor Style Guide](https://www.bioconductor.org/developers/how-to/coding-style/)

# Style Guides

- Each organization has a style guide on how code should be formatted that you should adhere to.

- When everyone on a project uses a consistent style, it makes code easier to read and understand, and it makes collaboration easier and faster.

- There are lots of style guides (see links in the Learning Objectives). This document contains the style guide for our class.

- This style guide is obviously opinionated, and others have their own thoughts (which is perfectly fine!). The important thing is consistency among collaborators.

- We will mostly follow the [tidyverse style guide](https://style.tidyverse.org/). Below I place some points of emphasis and note differences.

- I expect you to follow this style guide in all homeworks and assignments.

# File Names

- File and folder names should only have
    1. Letters
    2. Numbers
    3. Underscores (`_`).
    4. Possibly dashes `-`. But these are discouraged.
    
- In particular, never use spaces or periods in a file name.

- Capital letters are discouraged. You should work almost entirely with lower-case letters.

- Always begin a file name with a lower-case letter.

- Exceptions to this are:
    - Hidden files/folders begin with a period `.`
    - Standard/Required files, such as `NAMESPACE`, `README.md`, etc...
    
- R scripts should end in `.R` (not `.r`).

- R markdown files should end in `.Rmd`.

# Syntax


## Names

- Only use lower-case snake_case.

    - Good
        ```{r, eval = FALSE}
        red_apple
        ```
    
    - Bad
        ```{r, eval = FALSE}
        Red_apple
        red.apple
        redApple
        RedApple
        ```

- Variables should be nouns and functions should be verbs

- Never use single letters as variables/functions
    - Good:
        ```{r, eval = FALSE}
        num_sim <- 10
        ```
    - Bad
        ```{r}
        simulate <- 10 ## verb
        x <- 10 ## single letter
        ```

- **Exceptions**: Some letters are standard. Such as `n` for the sample size in `rnorm()`, `runif()`, etc...

## Commas

- Always put a space after a comma, not before (like English).
    - Good:
        ```{r, eval = FALSE}
        mat[1, ]
        ```
    - Bad:
        ```{r, eval = FALSE}
        mat[1 ,]
        mat[1 , ]
        mat[1,]
        ```

## Parentheses

- Don't put a space in or around parentheses for functions.
    - Good:
        ```{r, eval = FALSE}
        mean(x)
        ```
    - Bad:
        ```{r, eval = FALSE}
        mean (x)
        mean(x )
        ```
        
- Put spaces around parentheses for `if` statements, and `for` and `while` loops.
    - Good:
        ```{r, eval = FALSE}
        if (x) {
          
        }
        ```
    - Bad:
        ```{r, eval = FALSE}
        if(x){
          
        }
        ```

- Put a space only after `()` for function creations.
    - Good:
        ```{r, eval=FALSE}
        sim <- function(x) {
        }
        ```
    - Bad:
        ```{r, eval = FALSE}
        sim <- function (x) {
        }
        
        sim <- function(x){
        }
        ```
        
## Curley Braces

- Whenever you use curly braces `{}`, the opening brace should be the last character on a line, and the closing brace should be the first character on a line.
    - Good:
        ```{r, eval=FALSE}
        if (condition) {
          dostuff()
        }
        ```
    - Bad
        ```{r, eval=FALSE}
        if (condition)
        {
          dostuff()
        }
        
        if (condition) {
          dostuff() }
        ```
        
## `if-else`
        
- `else` statements should be on the same line as a closing brace.
    - Good:
        ```{r, eval=FALSE}
        if (condition) {
          
        } else if (condition2) {
          
        } else {
          
        }
        ```

- Only use `ifelse()` where vectorization is important. If `condition` should be length 1, then use full `if-else` statements.

- In a `if-then` statement, use `||` or `&&`, not `|` or `&`, since the latter two vectorize operations.
        
## Infix Characters

- An [**infix**](https://en.wikipedia.org/wiki/Infix_notation) operator is one where arguments on both sides of it are used in a function. The alternative is [**prefix**](https://en.wikipedia.org/wiki/Polish_notation) notation. Compare
    ```{r}
    5 + 10 ## infix notation
    `+`(5, 10) ## prefix notation
    ```

- Put spaces around all infix characters `==`, `+`, `-`, `*`, `/`, `^`, `|>`, etc...
    - Good:
        ```{r, eval = FALSE}
        x + 10
        ```
    - Bad:
        ```{r, eval=FALSE}
        x+10
        x+ 10
        x +10
        ```
        
- Exceptions: `::`, `:::`, `$`, `@`, `[`, `[[`, unary `-`, unary `+`, `:`, and `?`.
    - E.g. do `ggplot2::qplot()` or `-1`, not `ggplot2 :: qplot()` and `- 1`

## Code Length

- No lines should be greater than 80 characters.

- To get a vertical line displaying the code length, in R studio go to "Tools > Global Options... > Code > Display". Make sure "Show margin" is checked with "80" in the text box.

- If a function call/definition is too long, break up arguments on new line.
    ```{r, eval = FALSE}
    this <- is_a_very_long_function_call(
      that = "is",
      broken = "up",
      into = "many",
      indented = "lines",
      that = "are",
      easier = "to",
      read = NULL
    )
    ```

## Other things

- Always use `<-` for assignment, not `=`.

- Always use `"` for strings, not `'`.

- Always use `TRUE` or `FALSE`, not `T` or `F`
    - `T` and `F` are aliases for `TRUE` and `FALSE`, and so may be overwritten by the user, which is scary.
    
- Don't include [non-ASCII](https://en.wikipedia.org/wiki/ASCII#Character_set) characters in your code. 
    - ASCII characters are lower case letters (`a` through `z`), upper case letters (`A` through `Z`), digits (`0` through `9`), and common punctuation.
    - Including non-ASCII characters will give you a CRAN note.
    - Non-ASCII characters usually show up when you copy and paste from the web. E.g. the following look normal but are non-ASCII (and are all different): 
        - En Dash: "–"
        - Em Dash: "—"
        - Horizontal Bar: "―"
        - En Quad: " "
        - Em Quad: " "
        - En Space: " "
        - Em Space: " "
    - If you accidentally include such characters, you can find them with
        ```{r, eval = FALSE}
        tools::showNonASCIIfile()
        ```
    
# Functions

## Function Argument Length

- If you have a lot of arguments, indent the arguments on new lines.
    ```{r, eval = FALSE}
    run_me <- function(this,
                       is,
                       a = "lot",
                       of = "arguments",
                       that = "are longer than 80 characters") {
    }
    ```

## Function Length

- You should break up your functions into discrete tasks.
    - Reduces duplicating code, so less prone to bugs.
    - Allows you to think more modularly about tasks, which makes code easier to reason about.
    - Makes it easier to combine code in new ways.

- To force you to do this, make all functions be less than 50 lines. This is what [Bioconductor does](https://bioconductor.org/packages/devel/bioc/vignettes/BiocCheck/inst/doc/BiocCheck.html#function-length-checking).

## Explicit returns

- In R, the last value evaluated in a function will be implicitly returned. I think this is bad practice since it makes it harder to reason about what R is returning. So always include a `return()` statement. **Never** do
    ```{r, eval = FALSE}
    add_two <- function(x, y) {
      x + y
    }
    ```
    **Always** do
    ```{r, eval = FALSE}
    add_two <- function(x, y) {
      return(x + y)
    }
    ```

## Importing

- **Never** use the `@import` tag in a package to bring all of a package's exported functions into the `NAMESPACE`. This creates too much risk for name collision.

- In a package, never import functions, always type the package where the function came from. This makes it easier to reason about namespaces. **Never** do
    ```{r, eval=FALSE}
    #' @importFrom ggplot2 qplot
    plot_red <- function(x, y) {
      qplot(x, y, color = I("red"))
    }
    ```
    **Always** do
    ```{r, eval=FALSE}
    plot_red <- function(x, y) {
      ggplot2::qplot(x, y, color = I("red"))
    }
    ```
    
- **Exceptions**: 
    - You will have to import infix functions (surrounded by percent signs). Such as
        ```{r}
        #' @importFrom magrittr %>%
        #' @importFrom foreach %dopar%
        ```
    - There is a small performance penalty for using `::` (about 5 µs). So import a function if you are iterating it $\sim$ million times, and each iteration takes on the order of 1 ns.
    
## Order of Arguments

- Always place arguments with defaults after arguments without defaults.
- Good:
    ```{r, eval = FALSE}
    function(arg1, arg2, arg3 = NULL) {
      
    }
    ```
- Bad:
    ```{r, eval = FALSE}
    function(arg1, arg3 = NULL, arg2) {
      
    }
    ```

# `lintr`

- The `lintr` package will check many coding issues. Try running the following in the top directory of your package.
    ```{r, eval = FALSE}
    lintr::lint_package()
    ```

