---
title: "Web Scraping with rvest"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- [Overview of rvest](https://rvest.tidyverse.org/).
- [SelectorGadget](https://rvest.tidyverse.org/articles/selectorgadget.html).

# Data on the Web

- There are at least 4 ways people download data on the web:
    1. Click to download a csv/xls/txt file.
    2. Use a package that interacts with an API.
    3. Use an API directly.
    4. Scrape from directly from the HTML file.
    
- This lesson, we talk about how to do 4.

- Let's load the tidyverse:
    
    ```{r, message = FALSE}
    library(tidyverse)
    ```

# Scrape from HTML files with rvest

- To install SelectorGadget, drag this link to your bookmark bar on Chrome: [SelectorGadget](javascript:(function()%7Bvar%20s=document.createElement('div');s.innerHTML='Loading...';s.style.color='black';s.style.padding='20px';s.style.position='fixed';s.style.zIndex='9999';s.style.fontSize='3.0em';s.style.border='2px%20solid%20black';s.style.right='40px';s.style.top='40px';s.setAttribute('class','selector_gadget_loading');s.style.background='white';document.body.appendChild(s);s=document.createElement('script');s.setAttribute('type','text/javascript');s.setAttribute('src','https://dv0akt2986vzh.cloudfront.net/unstable/lib/selectorgadget.js');document.body.appendChild(s);%7D)();)

