---
title: "Simple Features and Graphing Maps"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.align  = "center",
                      fig.height = 3, fig.width = 4)
ggplot2::theme_set(ggplot2::theme_bw() + ggplot2::theme(strip.background = ggplot2::element_rect(fill = "white")))
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

# Learning Objectives

- Simple Features
    - <https://r-spatial.github.io/sf/articles/>
- Map Plotting
    - <https://ggplot2.tidyverse.org/reference/ggsf.html>

# Motivation

- For some reason, people are really impressed with data visualizations that include maps (e.g. of D.C., the U.S., the Earth, etc).

- Let's go over how to build simple map visualizations in R.

- For this lecture, we will try to visualize the 2020 presidential election results collected by the Upshot (<https://github.com/dcgerard/presidential-precinct-map-2020>).

# Simple Features

# Map Plotting
