---
title: "Plotting with Matplotlib and Seaborn"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 4, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Data visualization with seaborn and matplotlib
- Chapter 4 of [Python Data Science Handbook](https://jakevdp.github.io/PythonDataScienceHandbook/).

# Python Overview

| In R I Want | In Python I Use |
|-------------|-----------------|
| Base R      | numpy           |
| dplyr/tidyr | pandas          |
| ggplot2     | matplotlib/seaborn |

# Import Matplotlib and Seaborn, and Load Dataset

<mark style="background-color: lightblue">R</mark>
```{r}
library(ggplot2)
data("mpg")
```

All other code will be Python unless otherwise marked.

```{python}
import matplotlib.pyplot as plt
import seaborn as sns
mpg = r.mpg
```

# Show and clear plots.

- Use `plt.show()` to display a plot.

- Use `plt.clf()` to clear a figure when making a new plot.

# One Quantitative Variable: Histogram

- `sns.distplot()` makes a histogram/density plot.
    ```{python}
    sns.distplot(mpg.hwy)
    plt.show()
    ```
    
- Just obtain the histogram and choose number of bins:
    ```{python}
    sns.distplot(mpg.hwy, kde=False, bins=20)
    plt.show()
    ```

# One Categorical Variable: Barplot

- Use `sns.countplot()` to make a barplot to look at the distribution of a
  categorical variable:

    ```{python}
    sns.countplot(x='class', data=mpg)
    plt.show()
    ```

# One Quantitative Variable, One Categorical Variable: Boxplot

- Use `sns.boxplot()` to make boxplots:
    ```{python}
    sns.boxplot(x='class', y='hwy', data=mpg)
    plt.show()
    ```

- A boxenplot is a cool graphic that gives you more quantiles to look at:
    ```{python}
    sns.boxenplot(x='class', y='hwy', data=mpg)
    plt.show()
    ```

# Two Quantitative Variables: Scatterplot

- Use `sns.scatterplot()` to make a basic scatterplot.
    ```{python}
    sns.scatterplot(x='displ', y='hwy', data=mpg)
    plt.show()
    ```

## Lines/Smoothers

- Use `sns.regplot()` to make a scatterplot with a regression line or a loess smoother.

- Regression line with 95% Confidence interval
    ```{python}
    sns.regplot(x='displ', y='hwy', data=mpg)
    plt.show()
    ```
    
- Loess smoother with confidence interval removed.
    ```{python}
    sns.regplot(x='displ', y='hwy', data=mpg, lowess=True, ci='None')
    plt.show()
    ```

## Annotating by Third Variable

- Use the `hue` or `style` arguments to annotate by a categorical variable:
    ```{python}
    sns.scatterplot(x='displ', y='hwy', hue='class', data=mpg)
    plt.show()
    ```
    
    ```{python}
    sns.scatterplot(x='displ', y='hwy', style='class', data=mpg)
    plt.show()
    ```

- Use the `hue` or `size` arguments to annotate by a quantitative variable:

    ```{python}
    sns.scatterplot(x='cty', y='hwy', hue='displ', data=mpg)
    plt.show()
    ```
    
    ```{python}
    sns.scatterplot(x='cty', y='hwy', size='displ', data=mpg)
    plt.show()
    ```

# Two Categorical Variables: Mosaic Plot

- Usually, you should just show a table of proportions when you have two
  categorical variables.
  
- But if your boss wants a graphic, try a mosaic plot from the statsmodels package.
    ```{python, fig.height = 8, fig.width = 8, results = 'hide', fig.keep = 'all'}
    from statsmodels.graphics.mosaicplot import mosaic
    mosaic(data=mpg, index=['class', 'drv'])
    plt.show()
    ```
    
# Facets

- Use `sns.FacetGrid()` followed by the `map()` method to plot facets.
    ```{python}
    g = sns.FacetGrid(data=mpg, row='drv')
    g.map(sns.distplot, 'hwy', kde=False)
    plt.show()
    ```
    
# Labels

```{python, echo = FALSE}
plt.close()
```

- Assign plot to an object. Then use the `set_*()` methods to add labels.

    ```{python, fig.height=3, fig.width=6}
    scatter = sns.scatterplot(x='displ', y='hwy', data=mpg)
    scatter.set_xlabel('Displacement')
    scatter.set_ylabel('Highway')
    scatter.set_title('Highway versus Displacement')
    plt.show()
    ```

    
# Saving Figures

1. First, assign a figure to an object.

    ```{python, eval=FALSE}
    scatter = sns.scatterplot(x='displ', y='hwy', data=mpg)
    ```

2. Extract the figure. Assign this to an object.
    ```{python, eval=FALSE}
    fig = scatter.get_figure()
    ```

3. Save the figure.
    ```{python, eval=FALSE}
    fig.savefig('./scatter.pdf')
    ```

- You can do all of these steps using piping.
    ```{python, eval=FALSE}
    sns.scatterplot(x='displ', y='hwy', data=mpg) \
      .get_figure() \
      .savefig('./scatter.pdf')
    ```




