---
title: "Many Models"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---

```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3.5, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Use a list-column to run many models at once.
- Chapter 25 from [RDS](https://r4ds.had.co.nz/).

# Motivation

- The gapminder data
    ```{r, message = FALSE}
    library(tidyverse)
    library(broom)
    library(gapminder)
    data("gapminder")
    glimpse(gapminder)
    ```

- Suppose we want to look at how life expectancy has changed over time in 
  each country:
    ```{r}
    gapminder %>%
      ggplot(aes(x = year, y = lifeExp, group = country)) +
      geom_line(alpha = 1/3) +
      xlab("Year") +
      ylab("Life Expectancy")
    ```
    
- General trend is going up. But there are some countries where this doesn't
  happen. How do we quantify the trend? We can do this with one country
  using a linear model:
    ```{r}
    gapminder %>%
      filter(country == "United States") ->
      usdf
    
    ggplot(usdf, aes(x = year, y = lifeExp)) +
      geom_line() +
      geom_smooth(method = "lm", se = FALSE) +
      geom_line(alpha = 1/3) +
      xlab("Year") +
      ylab("Life Expectancy")
    
    us_lmout <- lm(lifeExp ~ year, data = usdf)
    tidy_uslm <- tidy(us_lmout)
    tidy_uslm
    ```
    
- So each year, the US has been increasing its life expectancy by 
  `r tidy_uslm$estimate[2]` years.
  
- How can we get these coefficient estimates for each country?

- For the exercises in this lesson plan, you should log the `pop` variable now:
    ```{r}
    gapminder %>%
      mutate(logpop = log2(pop)) ->
      gapminder
    ```


- **Exercise**: Create a line plot for year vs log-population for each country.
    ```{r, eval = FALSE, echo = FALSE}
    gapminder %>%
      ggplot(aes(x = year, y = pop, group = country)) +
      geom_line(alpha = 1/3) +
      xlab("Year") +
      ylab("Population") +
      scale_y_log10()
    ```


- **Exercise**: For China, fit a linear model for log-population on year. Make
  sure the assumptions of the linear model are fulfilled. Interpret the
  coefficients
    ```{r, eval = FALSE, echo = FALSE}
    gapminder %>%
      filter(country == "China") ->
      china_df
    
    ## some curvature caused by non-independence of years.
    china_df %>%
      ggplot(aes(x = year, y = logpop)) +
      geom_point() +
      geom_smooth(se = FALSE, method = "lm")
    
    china_lm <- lm(logpop ~ year, data = china_df)
    china_lm
    
    ## Each year, china grows by a factor of 2^0.0232 = 1.016.
    ```
  

# Nest

- The `nest()` function from the tidyr package will turn a *grouped* data frame
  into a data frame where each group is a row. All of the observations
  are placed in the `data` variable as a data frame. Let's look at an example.
    ```{r}
    gapminder %>%
      group_by(country, continent) %>%
      nest() ->
      nested_gap_df
    
    nested_gap_df
    ```
    
- The `data` variable is a list. Each element is a data frame that contains
  all of the observations in the group we created by `country`.
  
- Observations from Afghanistan
    ```{r}
    nested_gap_df$data[[1]]
    ```
  
- Observations from Albania
    ```{r}
    nested_gap_df$data[[2]]
    ```
  
- Notice that I used double brackets to extract the elements of `data` because
  it is a list.
  
- **Exercise**: Extract the United States data frame from `nested_gap_df`.
  This code should be *general* so that the location of the US in the data
  frame does not matter.
    ```{r, echo = FALSE, eval = FALSE}
    usind <- which(nested_gap_df$country == "United States")
    nested_gap_df$data[[usind]]
    ```
  
# Fit a model

- We can now use purrr to fit a model on all of the elements of `data`.
    ```{r}
    nested_gap_df %>%
      mutate(lmout = map(data, ~lm(lifeExp ~ year, data = .))) ->
      nested_gap_df
    ```

- Recall: the `data = .` argument says that we should place all of the
  elements from the `data` column in the `nested_gap_df` data frame
  as the `data` argument in `lm()`.

- Let's look at the results  
    ```{r}
    nested_gap_df
    ```
    
- Each element of the `lmout` column is the model output from fitting `lm()`
  to each data frame in the `data` column.
    ```{r}
    ## Afghanistan fit
    nested_gap_df$lmout[[1]]
    
    ## Albania fit
    nested_gap_df$lmout[[2]]
    ```
    
- **Exercise**: For each country, fit a linear model of log-population on
  year. Save this fit as the `lmpop` column in `nested_gap_df`.
    ```{r, eval = FALSE, echo = FALSE}
    nested_gap_df %>%
      mutate(lmpop = map(data, ~lm(logpop ~ year, data = .))) ->
      nested_gap_df
    ```
  
# Get model summaries

- Also use map to create columns of model summaries using `tidy()`, `augment()`,
  or `glance()`.
    ```{r}
    nested_gap_df %>%
      mutate(tidyout = map(lmout, ~tidy(., conf.int = TRUE)),
             augmentout = map(lmout, augment),
             glanceout = map(lmout, glance)) ->
      nested_gap_df
    ```

- We now have model summaries within all groups
    ```{r}
    ## Afghanistan summary
    nested_gap_df$tidyout[[1]]
    
    ## Albania summary
    nested_gap_df$tidyout[[2]]
    ```
    
- **Exercise**: Use `tidy()` to get model summaries of your linear model
  fits of log-population on year. Add these summaries as another column
  in `nested_gap_df`.
  
    ```{r, eval = FALSE, echo = FALSE}
    nested_gap_df %>%
      mutate(tidypop = map(lmpop, tidy)) ->
      nested_gap_df
    ```
  

# Unnest

- Use the `unnest()` function, followed by the `ungroup()` function, to 
  unnest a single list column.
    ```{r}
    unnest(nested_gap_df, tidyout) %>%
      ungroup() ->
      model_summary_df
    
    model_summary_df
    ```
    
- You do this so you can plot some cool things.
    ```{r}
    library(ggthemes)
    model_summary_df %>%
      filter(term == "year") %>%
      mutate(country = fct_reorder(country, estimate)) %>%
      ggplot(aes(x = country, y = estimate, color = continent)) +
      geom_point() +
      geom_segment(aes(x = country, xend = country, y = conf.low, yend = conf.high), 
                   color = "black", 
                   alpha = 1/3) +
      ylab("Estimate of Rate of Life Expectancy Increase") +
      xlab("Country") +
      theme(axis.text.x = element_blank(),
            axis.ticks.x = element_blank()) +
      scale_color_colorblind()
    ```
    
- You can check all of the models' qualities by looking at the residuals
    ```{r}
    unnest(nested_gap_df, augmentout) ->
      augment_df
    
    augment_df %>%
      ggplot(aes(x = year, y = .resid, group = country)) +
      geom_line(alpha = 1/3) +
      facet_wrap(.~continent) +
      geom_hline(yintercept = 0, lty = 2, col = "blue") +
      xlab("Residuals") +
      ylab("Year")
    ```
    
- Above, we see there is a lot of curvature we are not accounting for.


- **Exercise**: Unnest the `tidy()` output from the linear model fit of 
  log-population on year. Which countries have seen the greatest increase
  in population? Have any seen a decline?
    ```{r, eval = FALSE, echo = FALSE}
    unnest(nested_gap_df, tidypop) ->
      tpop
    
    tpop %>%
      filter(term == "year") %>%
      arrange(estimate) ->
      tpop
    
    tpop %>%
      select(country, estimate) %>%
      head()
    
    tpop %>%
      select(country, estimate) %>%
      tail()
    
    
    ## Bulgaria has had the slowest growth. Each year, its population stays about the same.
    ## Kuwait has had the fastest growth, each year having 5% growth (2 ^ 0.0699 = 1.05)
    ## We don't have any indication here of an overall population decline.
    ```



# `mtcars` exercise

Recall how we used the `split()` function in the iterators worksheet on the
`mtcars` dataset to get the regression coefficient estimates of a regression
of `mpg` on `wt`.

```{r}
data("mtcars")
mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .)) %>%
  map(~tidy(.)) %>%
  map_dbl(~.$estimate[2])
```

Redo this exercise using the `nest()`-`map()`-`unnest()` workflow we just
went through.
   
```{r, eval = FALSE, echo = FALSE}
mtcars %>%
  group_by(cyl) %>%
  nest() %>%
  mutate(lmout = map(data, ~lm(mpg ~ wt, data = .))) %>%
  mutate(tidyout = map(lmout, ~tidy(.))) %>%
  unnest(tidyout) %>%
  filter(term == "wt") %>%
  select(cyl, estimate)
```







  
  

