---
title: "The seaborn.objects Interface"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 4, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Data visualization with seaborn.objects.
- <https://seaborn.pydata.org/tutorial/objects_interface.html>

# Python Overview

| In R I Want | In Python I Use |
|-------------|-----------------|
| Base R      | numpy           |
| dplyr/tidyr | pandas          |
| ggplot2     | matplotlib/seaborn |

- I [previously](./06_matplotlib.html) taught how to use the basic seaborn interface. But in 2022 the author introduced an interface more similar to ggplot which seems to be the future of the package. So here is a lecture on the seaborns.objects interface.

- This interface is experimental, doesn't have all of the features you would need for data analysis, and will likely change. But I think it looks cool.

# Import Matplotlib and Seaborn, and Load Dataset

<mark style="background-color: lightblue">R</mark>
```{r}
library(ggplot2)
data("mpg")
```

All other code will be Python unless otherwise marked.

```{python}
import matplotlib.pyplot as plt # base plotting functionality
import seaborn as sns           # Original interface
import seaborn.objects as so    # ggplot2-like interface
mpg = r.mpg
```

# Basics

- Use `so.Plot()` to instantiate a `Plot` object and define asthetic mappings.
    - Just like `ggplot()`.

- Use the `.add()` method of the `Plot` object to add geometric objects and a statistical transformation.
    
- As in ggplot, each aesthetic mapping is followed by a statistical transformation before plotting. But unlike in ggplot, you need to specify this statistical transformation manually.
    - E.g. aggregating a categorical variable into a counts before plotting those counts on the y-axis.
    - E.g. binning a quantitative variable to make a histogram.
    
- You specify the statistical transformation as the second argument in `.add()`.
    
# One Quantitative Variable: Histogram

- Notice that we need to specify the `so.Hist()` statistical transformation to generate a histogram. 

- We use the `so.Bars()` geometric object to plot it after the statistical transformation.

    ```{python}
    pl = (
      so.Plot(mpg, x = "hwy")
      .add(so.Bars(), so.Hist(bins = 10))
    )
    pl.show()
    ```

# One Categorical Variable: Barplot

- We use the `so.Bar()` geometric object after the `so.Count()` statistical transformation.

    ```{python}
    pl = (
      so.Plot(mpg, x = "class")
      .add(so.Bar(), so.Count())
    )
    pl.show()
    ```

- `so.Bar()` (for categorical data) and `so.Bars()` (for quantitative data) seem to be only slightly different based on the defaults.

# Dodging Barplots

- If you are creating two barplots, annotated by color, you need to be explicit that the bars should dodge eachother with a `so.Dodge()` transformation.

    ```{python}
    pl = (
      so.Plot(mpg, x = "class", color = "drv")
      .add(so.Bar(), so.Count(), so.Dodge())
    )
    pl.show()
    ```


# One Quantitative Variable, One Categorical Variable: Boxplot

- This interface is currently (November 2022) missing boxplotting functions, so you need to use the old interface.

    ```{python}
    plt.clf()
    sns.boxplot(data = mpg, x = "class", y = "hwy")
    plt.show()
    plt.clf()
    ```

- I think this is the closest thing to a boxplot you can get right now:

    ```{python}
    pl = (
      so.Plot(mpg, x = "class", y = "hwy")
      .add(so.Dash(width = 0.4), so.Perc())
      .add(so.Range())
      .add(so.Range(), so.Perc([25, 75]), so.Shift(x=0.2))
      .add(so.Range(), so.Perc([25, 75]), so.Shift(x=-0.2))
    )
    pl.show()
    ```


    
# Two Quantitative Variables: Scatterplot

- Base scatterplot uses the `so.Dots()` geometric object:

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
    )
    pl.show()
    ```

- Use the `so.Jitter()` statistical transformation to make a jittered scatterplot.

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots(), so.Jitter(1))
    )
    pl.show()
    ```

- Use `so.Line()` (geometric object) and `so.PolyFit()` (statistical transformation) to add a smoother.

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit())
    )
    pl.show()
    ```
    
    - I don't think it does lowess or gam or splines yet, but just a polynomial, which is not optimal. You can control the order of the polynomial by the `order` argument.
    
    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit(order = 1))
    )
    pl.show()
    ```
    
- Annotate by a third variable by adding a color mapping:

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy", color = "drv")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit(order = 1))
    )
    pl.show()
    ```

# Faceting

- Facet by the `.facet()` method.

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .facet(row = "drv")
      .add(so.Dots())
    )
    pl.show()
    ```

# Customizing Look

- You can change the scaling using the `.scale()` method. E.g. here is a $\log_2$ transformation for the $x$-axis.

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit(order = 1))
      .scale(x = "log2")
    )
    pl.show()
    ```

- You can change the labels by `.label()`.

    ```{python}
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
      .label(x = "Displacement (L)", y = "Highway MPG")
    )
    pl.show()
    ```

- You can change the theme using `.theme()`. But it is a little verbose right now.

    ```{python}
    from seaborn import axes_style
    pl = (
      so.Plot(mpg, x = "displ", y = "hwy")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit(order = 1))
      .theme({**axes_style("whitegrid"), "grid.linestyle": ":"})
    )
    pl.show()
    ```

# Exercises

Consider the palmer penguins data, which you can load via

```{python}
penguins = sns.load_dataset("penguins")
penguins.info()
```

1. Make a visualization of bill length versus bill depth, annotated by species.

    ```{python, eval = FALSE, echo = FALSE}
    pl = (
      so.Plot(penguins, x = "bill_length_mm", y = "bill_depth_mm", color = "species")
      .add(so.Dots())
    )
    pl.show()
    ```

2. Add OLS lines to for each species to the **same** plot object you created in part 1 (don't rerun `so.Plot()`).

    ```{python, eval = FALSE, echo = FALSE}
    pl = pl.add(so.Line(), so.PolyFit(order=1))
    pl.show()
    ```

3. Use `pandas.cut()` to convert body mass into five equally spaced levels.

    ```{python, eval = FALSE, echo = FALSE}
    import numpy as np
    import pandas as pd
    penguins["bodycut"] = pd.cut(penguins.body_mass_g, bins = 5)
    ```

4. Facet your plot from part 2 by the above transformation. You will have to redo the object since we are using a different data frame here.

    ```{python, eval = FALSE, echo = FALSE, fig.height = 15}
    pl = (
      so.Plot(penguins, x = "bill_length_mm", y = "bill_depth_mm", color = "species")
      .add(so.Dots())
      .add(so.Line(), so.PolyFit(order=1))
      .facet(row = "bodycut")
    )
    pl.show()
    ```

5. Make a visualization for the number of each species in the dataset. Make sure you have good labels.

    ```{python, eval = FALSE, echo = FALSE}
    pl = (
      so.Plot(penguins, x = "species")
      .add(so.Bar(), so.Count())
      .label(x = "Species", y = "Count")
    )
    pl.show()
    ```
