---
title: "Simple Linear Regression"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives

- Simple Linear Regression
- ANOVA

# Pattern: Two quantitative variables are associated.

- Example: The big-bang theory suggests a linear relationship between 
  the recession velocity of a stellar object and its distance from Earth,
  with a y-intercept of 0. Measurements of recession velocity and distance
  were made on 24 extra-galactic nebulae.
    ```{r}
    data("case0701")
    bang <- case0701
    glimpse(bang)
    ```
    
- Observational Units: The nebulae.

- Population: All nebulae.

- Sample: The 24 nebulae that the researches measured.

- Variables: recession velocity (quantitative) and distance (quantitative).
  
- Pattern: Use a scatterplot to see if there is an association.

    ```{r}
    ggplot(bang, aes(x = Velocity, y = Distance)) +
      geom_point()
    ```

- Model: $Y_i = \beta_0 + \beta_1 X_i + noise$, where
    - $Y_i$ is the distance of nebula $i$.
    - $X_i$ is the velocity of nebula $i$.
    - $\beta_0$ is the intercept of the mean line.
    - $\beta_1$ is the slope of the mean line.

- Function: `lm()`

# Pattern: Means of three or more groups are different (one quantitative, one categorical).

- Example: [Dr. Spock](https://en.wikipedia.org/wiki/Benjamin_Spock) was on 
  trial for draft dodging. He claimed that his judge includes fewer women
  in his jury selection pool. They measured the proportion of women in 
  a jury selection in many trials for six judges as well as Spock's judge.
  
    ```{r}
    data("case0502")
    spock <- case0502
    glimpse(spock)
    ```
  
- Observational units: The trials in Boston.

- Population: All trials in Boston.

- Sample: The 46 trials the researchers collected data on.

- Variables: The percent of women in the jury selection pool (quantitative)
  and the judge of the trial (categorical).
  
- Pattern: Use a boxplot to see if the groups differ.

    ```{r}
    ggplot(spock, aes(x = Judge, y = Percent)) +
      geom_boxplot()
    ```

- Function: `lm()` or `aov()`.
