---
title: "SQL"
author: "David Gerard"
date: "`r Sys.Date()`"
output:  
  html_document:
    toc: true
    toc_depth: 4
urlcolor: "blue"
---

```{r createData, eval = FALSE, echo = FALSE}
## Code to write flights.duckdb
library(DBI)
library(duckdb)
library(nycflights13)
con <- dbConnect(duckdb(dbdir = "flights.duckdb", read_only = FALSE))
dbWriteTable(conn = con, name = "airlines", value = airlines)
dbWriteTable(conn = con, name = "airports", value = airports)
dbWriteTable(conn = con, name = "flights", value = flights)
dbWriteTable(conn = con, name = "planes", value = planes)
dbWriteTable(conn = con, name = "weather", value = weather)
dbDisconnect(con, shutdown = TRUE)
```


```{r setup, include=FALSE}
set.seed(1)
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Learn some SQL
- Interface SQL with R through the `{DBI}` and `{duckdb}` packages.
    - Introduction to DBI: <https://solutions.posit.co/connections/db/r-packages/dbi/>
    - DuckDB SQL Introduction: <https://duckdb.org/docs/sql/introduction>
    - DuckDB R API: <https://duckdb.org/docs/api/r.html>
- Write SQL code using the tidyverse and the `{dbplyr}` package.
    - `{dbplyr}` and SQL: <https://dbplyr.tidyverse.org/articles/sql.html>
    - R for Data Science Database Chapter: <https://r4ds.hadley.nz/databases.html>

# SQL Introduction and Motivation

- SQL is pronounced either "ess-kew-ell" or "sequel" (either is fine).

- SQL is a programming language which basically allows you to do a lot that dplyr does (filters, selects, joins, etc...). It is the industry standard for interacting with a relational database.

- You all are experts in data munging once a dataset has been downloaded.

- But, many datasets live in a relational database which is too huge to download onto your computer.

- A common task for a data scientist is to download a subset of this huge database so that they can explore it on their computer.

- SQL is how you download this subset. You choose which data frames ("tables" in SQL) to download, what subsets of these tables to download (by filtering rows), and whether to join tables together in this process.

- We will use R Studio in this lesson. But if you get really into SQL, a popular IDE is [DBeaver](https://dbeaver.io/), with a tutorial from DuckDB [here](https://duckdb.org/docs/guides/sql_editors/dbeaver).

# DuckDB

- There are a lots of [database management systems](https://en.wikipedia.org/wiki/Database#Database_management_system) (DBMS) that use SQL to connect to their databases.

- In this lecture, we will only consider [DuckDB](https://duckdb.org/) since it has no external dependencies and is pretty easy to install. Just do this in R:

    <mark style="background-color: lightblue">R</mark>
    
    ``` r
    install.packages("duckdb")
    ```

- The most popular DBMS is probably [SQLite](https://www.sqlite.org/index.html). Other popular ones are [MySQL](https://www.mysql.com/) and [PostgreSQL](https://www.postgresql.org/).

- DuckDB is a specific SQL backend. The R package that is used to connect to any SQL backend is `{DBI}`, and so we will load that one too.

    <mark style="background-color: lightblue">R</mark>
    ```{r, message = FALSE}
    library(DBI)
    library(duckdb)
    ```
    
- We will compare to the tidyverse:

    <mark style="background-color: lightblue">R</mark>
    ```{r, message = FALSE}
    library(nycflights13)
    library(tidyverse)
    ```

- Let's download a duck database that I created. Put this in a location you can get to: <https://data-science-master.github.io/lectures/data/flights.duckdb>

- Use `duckdb()` and `DBI::dbConnect()` to create a connection to "flights.duckdb".

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    con <- dbConnect(duckdb(dbdir = "../data/flights.duckdb", read_only = TRUE))
    ```

- A basic SQL code chunk looks like this (put SQL code between the chunks):

    ```{r, echo = FALSE, comment = ""}
    codechunk <- "```{sql, connection=con}\n\n```"
    writeLines(codechunk)
    ```

# Basic SQL

## General

- Case does not matter (i.e. `select` is the same as `SELECT` is the same as `SeLeCt`), but it is standard to have all statements be in UPPERCASE (i.e. `SELECT`).

- The statements below **must** be in the following order: `SELECT`, `FROM`, `WHERE`, `GROUP BY`, `ORDER BY`.

- New lines and white space don't matter. But it is common to put those five commands above on new lines.

- Character values must be in single quotes.

- You can use invalid variable names by putting them in double quotes (same as using backticks in R).
    - Some folks always use double quotes because it is not always clear what is an invalid variable name in the database management system. This is what I do below.

- Comments in SQL have two hyphens `--`.

- Make sure to put a semicolon `;` at the end of a SQL statement. This will allow you to have multiple SQL queries in one chunk.

## Showing Tables

- The `SHOW TABLES` command can be used to get a list of all of the tables

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SHOW TABLES;
    ```

- The `DESCRIBE` command can be used to show tables and the variables.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    DESCRIBE;
    ```

## Select Specific Columns

- To get the columns from a data frame, use the `SELECT` command.

- The syntax is like this

    <mark style="background-color: #e24e2f">SQL</mark>
    
    ``` sql
    SELECT <column1>, <column2>, <column3> 
    FROM <mytable>;
    ```
    
- Let's use this to get the `tailnum`, `year`, and `model` variables from the `planes` table.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "tailnum", "year", "model" 
    FROM planes;
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    planes %>%
      select(tailnum, year, model)
    ```

- You select every column by using `*`:

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT * 
    FROM planes;
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    planes
    ```

- There is no equivalent for excluding columns (like `dplyr::select(-year)`). You just select the ones you want.

## Filter rows

- You use the `WHERE` command in SQL to filter by rows.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "distance", "origin", "dest" 
    FROM flights
    WHERE "distance" < 50;
    ```
    
- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, distance, origin, dest) %>%
      filter(distance < 50)
    ```

    
- To test for equality, you just use one equals sign.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "month" 
    FROM flights
    WHERE "month" = 12;
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, month) %>%
      filter(month == 12)
    ```


- For characters you **must use single quotes**, not double.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "origin" 
    FROM flights
    WHERE "origin" = 'JFK';
    ```
    
- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, origin) %>%
      filter(origin == "JFK")
    ```


- You can select multiple criteria using the `AND` command

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "origin", "dest" 
    FROM flights
    WHERE "origin" = 'JFK' AND "dest" = 'CMH';
    ```

- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, origin, dest) %>%
      filter(origin == "JFK", dest == "CMH")
    ```


- You can use the `OR` logical operator too. Just put parentheses around your desired order of operations.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "origin", "dest" 
    FROM flights
    WHERE ("origin" = 'JFK' OR "origin" = 'LGA') AND dest = 'CMH';
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, origin, dest) %>%
      filter(origin == "JFK" | origin == "LGA", dest == "CMH")
    ```

- Missing data is `NULL` in SQL (instead of `NA`). We can remove them by the special command:

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "dep_delay" 
    FROM flights
    WHERE "dep_delay" IS NOT NULL;
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, dep_delay) %>%
      filter(!is.na(dep_delay))
    ```

- Just use `IS` if you want only the missing data observations

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "dep_delay" 
    FROM flights
    WHERE "dep_delay" IS NULL;
    ```

## Arranging Rows

- Use `ORDER BY` to rearrange the rows (let's remove missing values so we can see the ordering)

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "dep_delay" 
    FROM flights
    WHERE "dep_delay" IS NOT NULL
    ORDER BY "dep_delay";
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, dep_delay) %>%
      filter(!is.na(dep_delay)) %>%
      arrange(dep_delay)
    ```
    
- Use `DESC` after the variable to arrange in descending order

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "dep_delay" 
    FROM flights
    WHERE "dep_delay" IS NOT NULL
    ORDER BY "dep_delay" DESC;
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, dep_delay) %>%
      filter(!is.na(dep_delay)) %>%
      arrange(desc(dep_delay))
    ```

- You break ties by adding more variables in the `ORDER BY` statement

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "origin", "dep_delay" 
    FROM flights
    WHERE "dep_delay" IS NOT NULL
    ORDER BY "origin" desc, "dep_delay";
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, origin, dep_delay) %>%
      filter(!is.na(dep_delay)) %>%
      arrange(desc(origin), dep_delay)
    ```

## Mutate

- In SQL, you mutate variables while you SELECT. You use `AS` to specify what the new variable is called (choosing a variable name is called "aliasing" in SQL).

    <mark style="background-color: #e24e2f">SQL</mark>
    
    ``` sql
    SELECT <expression> AS <myvariable> 
    FROM <mytable>;
    ```

- Let's calculate average speed from the `flights` table. We'll also keep the flight number, distance, and air time variables.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "distance" / "air_time" AS "speed", "distance", "air_time" 
    FROM flights;
    ```

- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, distance, air_time) %>%
      mutate(speed = distance / air_time)
    ```
    
- Various transformation functions also exist:
    - `LN()`: Natural log transformation.
    - `EXP()`: Exponentiation.
    - `SQRT()`: Square root.
    - `POW()`: Power transformation. 
        - `POW(2.0, x)` would be $2^x$
        - `POW(x, 2.0)` would be $x^2$.

## Group Summaries

- SQL has a few summary functions (SQL calls these "[Aggregates](https://duckdb.org/docs/sql/aggregates)")
    - `COUNT()`: Count the number of rows.
    - `AVG()`: Calculate average.
    - `MEDIAN()`: Median (not standard across all DBMS's).
    - `SUM()`: Summation.
    - `MIN()`: Minimum.
    - `MAX()`: Maximum.
    - `STDDEV()`: Standard deviation.
    - `VARIANCE()`: Variance
    
- By default, all missing data are ignored (like setting `na.rm = TRUE`).
    
- These are calculated in a `SELECT` command

- Let's calculate the average departue delay

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT AVG("dep_delay") AS "dep_delay" 
    FROM flights;
    ```

- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      summarize(dep_delay = mean(dep_delay, na.rm = TRUE))
    ```
    
- Use the `GROUP BY` command to calculate group summaries.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "origin", AVG("dep_delay")
    FROM flights
    GROUP BY "origin";
    ```

- R equivalent

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(origin, dep_delay) %>%
      group_by(origin) %>%
      summarize(dep_delay = mean(dep_delay, na.rm = TRUE))
    ```

- You can get distinct rows by using the prefix `DISTINCT` in a `SELECT` statement. E.g. the following will pick up all unique origins

    ```{sql, connection=con}
    SELECT DISTINCT "origin" 
    FROM flights;
    ```

## Recoding

- Use the following `CASE`-`WHEN` syntax to recode values

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "origin", CASE
        WHEN ("origin" = 'JFK') THEN 'John F. Kennedy'
        WHEN ("origin" = 'LGA') THEN 'LaGaurdia'
        WHEN ("origin" = 'EWR') THEN 'Newark Liberty'
        END AS "olong"
    FROM flights;
    ```

- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, origin) %>%
      mutate(olong = case_when(
        origin == "JFK" ~ "John F. Kennedy",
        origin == "LGA" ~ "LaGuardia",
        origin == "EWR" ~ "Newark Liberty")
      )
    
    ## or
    flights %>%
      select(flight, origin) %>%
      mutate(olong = recode(
        origin,
        "JFK" = "John F. Kennedy",
        "LGA" = "LaGuardia",
        "EWR" = "Newark Liberty")
      )
    ```
    
- You can also use `CASE`-`WHEN` to recode based on other logical operations

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    SELECT "flight", "air_time", CASE
        WHEN ("air_time" > 2500) THEN 'Long'
        WHEN ("air_time" <= 2500) THEN 'Short'
        END AS "qual_dist"
    FROM flights;
    ```
    
- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r, eval = FALSE}
    flights %>%
      select(flight, air_time) %>%
      mutate(qual_dist = case_when(
        air_time > 2500 ~ "Long",
        air_time <= 2500 ~ "Short")
      )
    
    ## or
    flights %>%
      select(flight, air_time) %>%
      mutate(qual_dist = if_else(air_time > 2500, "Long", "Short"))
    ```
    
## Joining

- For joining, in the `SELECT` call, you write out all of the columns in **both** tables that you are joining.

- If there are shared column names, you need to distinguish between the two via `table1."var"` or `table2."var"` etc...

- Use `LEFT JOIN` to declare a left join, and `ON` to declare the keys.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    -- flight is from the flights table
    -- type is from the planes table
    -- both tables have a tailnum column, so we need to tell them apart
    -- if you list both tailnums in SELECT, you'll get two tailnum columns
    SELECT "flight", flights."tailnum", "type" 
    FROM flights
    JOIN planes
    ON flights."tailnum" = planes."tailnum";
    ```

- R equivalent:

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    planes %>%
      select(tailnum, type) ->
      planes2
    flights %>%
      select(flight, tailnum) %>%
      left_join(planes2, by = "tailnum")
    ```

- The other joins are:
    - `RIGHT JOIN`
    - `INNER JOIN`
    - `FULL OUTER JOIN`

## Creating some Tables

- You can create tables using SQL.

- Let's create a new, temporary, connection that we can put some example tables in.

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    tmpcon <- dbConnect(duckdb())
    ```

- You can use `CREATE TABLE` to declare new tables and `INSERT INTO` to insert new rows into that table. 
    - Note that I am putting a semicolon "`;`" after each SQL call in the same code chunk.

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=tmpcon}
    CREATE TABLE table4a (
        "country" VARCHAR,
        "1999"   INTEGER,
        "2000"   INTEGER
    );
    
    INSERT INTO table4a ("country", "1999", "2000") 
        VALUES
            ('Afghanistan',    745,   2666),
            ('Brazil'     ,  37737,  80488),
            ('China'      , 212258, 213766);
    ```
    
- But `DBI::dbWriteTable()` will add a table to a connection from R.

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    data("table1", package = "tidyr")
    dbWriteTable(conn = tmpcon, name = "table1", value = table1)
    data("table2", package = "tidyr")
    dbWriteTable(conn = tmpcon, name = "table2", value = table2)
    data("table3", package = "tidyr")
    dbWriteTable(conn = tmpcon, name = "table3", value = table3)
    data("table4b", package = "tidyr")
    dbWriteTable(conn = tmpcon, name = "table4b", value = table4b)
    data("table5", package = "tidyr")
    dbWriteTable(conn = tmpcon, name = "table5", value = table5)
    ```

- Here they are:

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=tmpcon}
    DESCRIBE;
    ```

## Tidyr stuff

- Recall the four main verbs of tidying: gathering, spreading, separating, and uniting.

- These are very rarely used in SQL. 

- SQL is used mostly to query a subset of data. You would then load in that data using more advanced methods (R, Python, Excel, etc).

- It's still possible to spread and gather (but not separate and unite), but a huge pain and folks don't typically do it.

## Writing Tables to a CSV File

- You can use `COPY` `TO` to write the outputs of a SQL query to a CSV file.

- The syntax for this is

    <mark style="background-color: #e24e2f">SQL</mark>
    
    ``` sql
    COPY
    (
    Your SQL query goes here
    )
    TO 'myfile.csv' (HEADER, DELIMITER ',');
    ```

- Let's write some group summaries to a CSV file

    <mark style="background-color: #e24e2f">SQL</mark>
    ```{sql, connection=con}
    COPY
    (
    SELECT "origin", AVG("dep_delay")
    FROM flights
    GROUP BY "origin"
    )
    TO 'summaries.csv' (HEADER, DELIMITER ',');
    ```

- This is what the resulting file looks like:

    ```{r, echo = FALSE}
    writeLines(readLines("./summaries.csv"))
    ```

# Calling SQL from R

- Use `DBI::dbGetQuery()` to run SQL code in R and obtain the result as a data frame.

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    planes2 <- DBI::dbGetQuery(conn = con, statement = "SELECT * FROM planes")
    glimpse(planes2)
    ```
    
- Often, folks have SQL written in a separate file (that ends in ".sql").

- E.g., here is what is in "query.sql":

    ```{r, echo = FALSE, comment=""}
    writeLines(readLines("query.sql"))
    ```

- If you want to load the results of a SQL query in R, saved in "query.sql", do

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    mydf <- DBI::dbGetQuery(conn = con, statement = read_file("query.sql"))
    glimpse(mydf)
    ```

# Use R to generate SQL with `{dbplyr}`

- `{dbplyr}` allows you to use `{dplyr}` code on a SQL backend.

- It will even generate SQL code for you, so you can use it as a way to learn SQL.

- Let's load it in:

    <mark style="background-color: lightblue">R</mark>
    ```{r, message=FALSE}
    library(dbplyr)
    ```

- You retrieve a table from a SQL database using `tbl()`

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    flights2 <- tbl(src = con, "flights")
    ```

- Now you can use your tidyverse code on the flights table

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    flights2 %>%
      select(flight, origin, dest, dep_delay) %>%
      filter(origin == "JFK", dest == "CMH") %>%
      summarize(dep_delay = mean(dep_delay, na.rm = TRUE))
    ```
    
- You execute the query by using `collect()` at the end.

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    flights2 %>%
      select(flight, origin, dest, dep_delay) %>%
      filter(origin == "JFK", dest == "CMH") %>%
      summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
      collect()
    ```

- To see the SQL code that was generated, use `show_query()` at the end (this is a good way to learn SQL).

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    flights2 %>%
      select(flight, origin, dest, dep_delay) %>%
      filter(origin == "JFK", dest == "CMH") %>%
      summarize(dep_delay = mean(dep_delay, na.rm = TRUE)) %>%
      show_query()
    ```

# Closing the Connection

- After you are done with SQL, you should close down your connection:

    <mark style="background-color: lightblue">R</mark>
    ```{r}
    DBI::dbDisconnect(con, shutdown = TRUE)
    DBI::dbDisconnect(tmpcon, shutdown = TRUE)
    ```

# Exercises

```{r, eval=FALSE, echo=FALSE}
## Code to write starwars.duckdb
library(DBI)
library(duckdb)
library(starwarsdb)
con <- dbConnect(duckdb(dbdir = "starwars.duckdb", read_only = FALSE))
dbWriteTable(conn = con, name = "films", value = films)
dbWriteTable(conn = con, name = "people", value = people)
dbWriteTable(conn = con, name = "planets", value = planets)
dbWriteTable(conn = con, name = "species", value = species)
dbWriteTable(conn = con, name = "vehicles", value = vehicles)
dbWriteTable(conn = con, name = "pilots", value = pilots)
dbWriteTable(conn = con, name = "films_people", value = films_people)
dbWriteTable(conn = con, name = "films_planets", value = films_planets)
dbWriteTable(conn = con, name = "films_vehicles", value = films_vehicles)
dbDisconnect(con, shutdown = TRUE)
```

The file [starwars.duckdb](https://data-science-master.github.io/lectures/data/starwars.duckdb) contains the Star Wars database from the [`{starwarsdb}`](https://cran.r-project.org/package=starwarsdb) R package.

1. Download these data and open a connection to the database.

    ```{r, echo = FALSE, eval=TRUE}
    library(DBI)
    library(duckdb)
    starcon <- dbConnect(duckdb(dbdir = "../data/starwars.duckdb", read_only = TRUE))
    ```

2. Print out a summary of the tables in this database.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    DESCRIBE;
    ```

3. Select the entire `people` table.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT *
    FROM people;
    ```

3. Select just the `name`, `height`, `mass`, and `species` variables from the `people` table.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT "name", "height", "mass", "species"
    FROM people;
    ```

4. Add to the above query by selecting only the humans and droids.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT "name", "height", "mass", "species"
    FROM people
    WHERE "species" = 'Human' OR "species" = 'Droid';
    ```

5. Remove the individuals with missing `mass` data from the above query.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT "name", "height", "mass", "species"
    FROM people
    WHERE ("species" = 'Human' OR "species" = 'Droid') AND "mass" IS NOT NULL;
    ```

6. Modify the above query to calculate the average height and mass for humans and droids.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT "species", AVG("height") AS "height", AVG("mass") AS "mass"
    FROM people
    WHERE ("species" = 'Human' OR "species" = 'Droid') AND "mass" IS NOT NULL
    GROUP BY "species";
    ```

7. Make sure that Droids are in the first row and Humans are in the second row in the above summary output.

    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    SELECT "species", AVG("height") AS "height", AVG("mass") AS "mass"
    FROM people
    WHERE ("species" = 'Human' OR "species" = 'Droid') AND "mass" IS NOT NULL
    GROUP BY "species"
    ORDER BY "species";
    ```

8. Here is the summary of the keys for the database from the [`{starwarsdb}` GitHub page](https://github.com/gadenbuie/starwarsdb).

    ![](./README-starwars-data-model-1.svg)\ 

    Select all films with characters whose homeworld is Kamino.
    
    ```{sql, connection=starcon, echo=FALSE, eval=FALSE}
    FROM (
    SELECT DISTINCT title AS title
    FROM people, films_people,
    WHERE people."homeworld" = 'Kamino' AND people."name" = films_people."character"
    GROUP BY "title"
    ) LHS
    INNER JOIN films
    ON LHS."title" = films."title"
    ```





